\documentclass[11pt]{article}
\usepackage{report}

\title{Nonlinear Band-Edge Sensing in Periodic Elastic Waveguides:\\
Drift-Resistant Acoustic Thermometry for Battery Safety Monitoring}
\author{Kazi Tahsin Mahmood}
\affil{Department of Mechanical Engineering, Wayne State University}
\date{}

\begin{document}

\maketitle

\section{Intellectual Merit}

\paragraph{Challenge.}
Lithium-ion battery thermal runaway is a persistent safety risk in transportation, grid storage, and defense. Interior hot-spot temperatures precede external warning signs by minutes to hours, yet all deployed monitoring technologies require penetrative wiring through the cell housing or measure surface temperature only. Non-contact ultrasonic time-of-flight (TOF) has shown promise as a non-invasive thermometry channel~\cite{Cheng2024}, but TOF is a single-channel, amplitude-agnostic observable that drifts with coupling-layer condition, electrolyte gas evolution, and drive-level variation, limiting practical accuracy to $\pm$1--2$^\circ$C even under controlled conditions. Overcoming these limitations calls for exploiting nonlinear periodic-structure physics to structurally amplify sensitivity, combined with phase-sensitive acoustic instrumentation to suppress drift.

\paragraph{Why This is New and Why Now.}
Linear ultrasonic battery thermometry exists~\cite{Cheng2024}, but exploiting band-edge amplification of second-harmonic generation as a temperature discriminator has not been attempted. The critical novelty is treating the periodic elastic structure not as a passive wave conduit but as an active sensing element whose nonlinear frequency response encodes temperature through the $1/\Delta^2$ amplification mechanism: near the band edge the group velocity approaches zero (slow-wave condition), wave energy accumulates in the unit cells, and small temperature-induced stiffness shifts produce disproportionately large changes in the harmonic ratio $A_{2f}/A_f$, far exceeding linear-TOF sensitivity. This approach is timely for three reasons. First, the theoretical foundations for slow-wave nonlinear enhancement in elastic lattices are now sufficiently mature~\cite{Lazarov2007} to support a reliable analytical-to-experimental translation. Second, precision CNC machining enables the asymmetric stub geometries required for Duffing-type nonlinearity at commercially relevant tolerances. Third, the rapid growth of lithium-ion battery deployment (in electric vehicles, stationary grid storage, and defense systems) creates urgent, unmet demand for non-invasive interior temperature monitoring at scales where the $\pm$1--2$^\circ$C accuracy ceiling of existing TOF methods is no longer acceptable. NewFoS targets acoustic sensing solutions for energy and transportation industries; this project delivers a new sensing principle (nonlinear band-edge thermometry) that closes an interior battery temperature monitoring gap no existing NewFoS linear-acoustic method has addressed.

\paragraph{PI Background.}
The PI's research expertise spans nonlinear vibration, acoustics, and periodic elastic structure theory, precisely the intersection this project requires. Prior work demonstrates phase-sensitive lock-in detection of nonlinear harmonic states in piezo-driven acoustic waveguides~\cite{Mahmood2022} and a Bloch-sphere dispersion formalism for one-dimensional elastic lattices covering band-edge analysis and symmetry-governed mode hybridization~\cite{Mahmood2026a}. These contributions establish working expertise in harmonic balance, Floquet-Bloch dispersion, multiple-scales perturbation, and the experimental measurement of harmonic generation in acoustic systems. Critically, the PI is among the few researchers who combine the nonlinear dynamics theory \emph{and} the acoustic instrumentation capability needed to bridge slow-wave amplification theory and practical thermometry, a combination that makes this project both uniquely feasible and difficult to replicate elsewhere.


\section{Research Plan}

\paragraph{Tasks.}
\begin{itemize}[noitemsep]
  \item \textbf{Task~1: Analytical model (months 1--4):} \emph{Problem:} No predictive model links the unit-cell Duffing coefficient, slow-wave detuning $\Delta$, and temperature sensitivity of $A_{2f}/A_f$; without it the optimal drive frequency cannot be set a priori. \emph{Solution:} Harmonic balance at $f_0$ and $2f_0$ applied to a Duffing diatomic chain yields $A_{2f}/A_f \propto \delta_\mathrm{NL}/\Delta^2$; multiple-scales analysis then provides the solvability condition bounding the valid drive-amplitude range and shows that differentiation with respect to temperature (aluminum modulus: $-40$\,ppm/$^\circ$C) gives $\partial(A_{2f}/A_f)/\partial T \propto 2\delta_\mathrm{NL}|\partial f_\mathrm{edge}/\partial T|/\Delta^3$, a $\Delta^{-3}$ sensitivity amplification steeper than the signal itself. The optimal detuning $\Delta^*$ is identified as the maximizer of this expression subject to the perturbation validity bound. The resulting sensitivity map, parameterized by $\delta_\mathrm{NL}$ and $\Delta$, directly informs the stub geometry ranges explored in Task~2.

  \item \textbf{Task~2: FE simulation (months 3--8):} \emph{Problem:} The 1D chain model neglects plate boundary conditions, two-dimensional wave spreading, and higher-order modes that can contaminate the $2f$ channel; these effects must be resolved before fabrication. \emph{Solution:} A 30--40\,cm 6061-T6 aluminum plate waveguide (8--12 unit cells, CNC-asymmetric stubs) is implemented in a finite element environment with temperature-dependent elastic constants; a parametric sweep of stub height, width, and asymmetry ratio maps $\partial(A_{2f}/A_f)/\partial T$ while $3f$/$4f$ monitoring ensures a clean harmonic operating regime. Stub geometry is additionally constrained to $\pm$0.1\,mm machining tolerance throughout the sweep, ensuring that the sensitivity maximum identified computationally is reproducible during CNC fabrication. Agreement within 10\% with the Task~1 closed-form expression validates the analytical model and fixes the final geometry.

  \item \textbf{Task~3: Fabrication and experiment (months 6--12):} \emph{Problem:} Amplifier source harmonics at $2f_0$ and coupling-layer drift can mimic the structural temperature signal; genuine drift resistance must be demonstrated against controlled disturbances. \emph{Solution:} A transmit bandpass filter suppresses source harmonics; $A_{2f}/A_f$ is extracted via coherent FFT averaging over 20 tone-bursts per step as the specimen is stepped through 20--100$^\circ$C (5$^\circ$C increments) with drive voltage sinusoidally modulated $\pm$20\% to emulate coupling variation. A symmetric-stub control specimen eliminates bond and amplifier nonlinearity as confounds. Sensitivity (dB/$^\circ$C) and drift-induced temperature error are reported with 95\% confidence intervals from ten repeated temperature cycles, providing a statistically robust comparison between the nonlinear waveguide and the linear TOF control.
\end{itemize}

\paragraph{Milestones, risk, and timeline.}
\textbf{M1 (month 4):} Closed-form sensitivity model derived; draft manuscript submitted. \textbf{M2 (month 8):} Finite element sensitivity map complete; results presented at a conference. \textbf{M3 (month 12):} Plate specimen fabricated and tested; sensitivity and drift-rejection data collected; journal manuscript submitted. The primary risk (stub nonlinearity too small to produce a measurable second-harmonic ratio) is mitigated by (a)~a symmetric-stub control for common-mode cancellation, (b)~stub geometry optimization in Task~2 before machining, and (c)~input harmonic filters at the transmitter to suppress source spurious harmonics. \textbf{Timeline:} Sept--Dec~2026 analytical modeling; Jan--Apr~2027 finite element simulation and plate fabrication; May--Aug~2027 experiments, data analysis, and manuscript writing.


\section{Broader Impact}

\paragraph{Technological.}
This project establishes a new acoustic sensing principle in which temperature sensitivity is structurally amplified by slow-wave band-edge dynamics in a nonlinear periodic structure, rather than derived from a direct wave-speed measurement. The modality generalizes from temperature to mechanical strain, state-of-charge, and electrolyte degradation by redesigning the unit-cell geometry to shift the band edge in response to the target quantity. Successful demonstration opens follow-on funding pathways through NSF CMMI, DoE Vehicle Technologies Office, and DoD ONR, and a Year-2 effort will validate the sensor against commercial Li-ion cells under charge-discharge cycling.

\paragraph{Workforce and education.}
This project will develop the PI's expertise at the intersection of nonlinear vibration theory, periodic elastic structure design, acoustic instrumentation, and applied sensing, a rare and commercially valuable combination spanning the aerospace, automotive, and energy-storage industries. Undergraduate research opportunities will be targeted at NewFoS partner Spelman College. All finite element models, analysis scripts, fabrication drawings, and raw calibration datasets will be released as open-source resources through the NewFoS community data portal.

{%
\renewcommand*{\bibfont}{\footnotesize}
\setlength{\bibitemsep}{2pt}
\setlength{\bibnamesep}{1pt}
\printbibliography
}

\end{document}
